RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-03-14 inst/doc/pubs/manuscript.Rmd
working on manuscript

```{r}
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

@@

```{r}
nex <- add_trees(geospiza$phy)
nex <- add_characters(geospiza$dat, nex)
```

@@

```{r}
data(primates)
nex <- add_trees(primates$phy, nex)
nex <- add_characters(primates$dat, nex)
```

@@

```{r}
data(bird.orders)
nex <- add_trees(bird.orders)
nex <- add_basic_meta(nex,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

@@

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.

- Utilities to add common metadata

@@

Writing annotations on otus.  Annotations on nodes.

```{r}
nex <- taxize_nexml(nex, "NCBI")
```

Temporal, spatial, taxonomic coverage for dryad/knb?

@@

To illustrate metadata access, we begin by reading in a nexml file.

>>>>

```{r}
library(geiger)
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

@@

```{r}
geiger_nex <- add_trees(geospiza$phy)
geiger_nex <- add_characters(geospiza$dat, geiger_nex)
```

@@

```{r}
data(primates)
geiger_nex <- add_trees(primates$phy, geiger_nex)
geiger_nex <- add_characters(primates$dat, geiger_nex)
```

@@

```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(birds,
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

@@

Citations can also be added using R's `bibentry` type, which allows
a user to add citations to R packages,

```{r}
birds <- add_basic_meta(birds, citation = citation("ape"))
```

or to papers by simply using the object's DOI:

```{r}
library("knitcitations")
geiger_nex <- add_basic_meta(geiger_nex, citation=cite("10.2307/2408428"))
```

(Which is the paper on which the `geospiza` phylogeny from the geiger dataset is based)

`RNeXML` also provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.

```{r}
birds <- taxize_nexml(birds, "NCBI")
```

### Custom metadata extensions

So far our examples have relied on functions that have a built-in understanding of the 
kind of metadata being provided, and can thus generate the appropriate metadata with 
little further guidence.  In order to be machine-readable and understandable, 
NeXML requires all metadata annotations follow the RDFa format which explicitly declares 
the meaning of the annotation in a way a computer can understand.  For instance, the 
title we gave to the `bird.orders` phylogeny appears in the NeXML as:

```xml
<meta id="m2" property="dc:title" datatype="xsd:string" content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
```

We see the title appears as the `content` of this `meta` element.  The `id` attribute
provides us with a unique way to refer to this meta element in this document.  The `datatype`
tells us that the content is a text string (as opposed to a numerical or boolean value).  
The `property` is the attribute that most interests us here; it tells us that the content 
is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:

```{r}
prefixes <- get_namespaces(birds)
prefixes["dc"]
```

This URL links us to a webpage providing both a human and machine-readable version of precisely
what the term `title` means, showing us that it refers to `title` in the sense of a creative work, 
rather than the title of a person, a legal title, the verb title, etc.  This precision can allow
us to perform precise computer reasoning on metadata, without relying on heuristics or assumptions,
as we illustrate later.  For the moment though, the need to refer to an explicit controlled vocabulary (or
ontology) with a url, saying "dc:title" rather than just saying the property is "title", is something
of a burden in creating metadata.  The previous examples involve functions that already know what 
kind of metadata is being provided (a title, a data creator, etc), and so can provide the appropriate
vocabulary automatically.  The main advantage of linking to an external vocabulary, however, is the
ability to express concepts that are not already "built in" to the data model.  `RNeXML` provides 
tools to generate arbitrary metadata annotations given an external vocabulary, such as we illustrate
here.

From the list of namespaces in `prefixes`, we can follow links a handful of established vocabularies
and find terms that are not included in `add_basic_meta`.  For instance, in the `prism` prefix,

```{r}
prefixes["prism"]
```

we see that there is a term for `modificationDate` for a file.  We create a `meta` element 
containing this annotation using the `meta` function:

```{r}
modified <- meta(property = "prism:modificationDate", content = "2013-10-04")
```

We can add this annotation to our existing `birds` nexml file using the `add_meta` function.
Because we do not specify a level, it is added to the root node, refering to the nexml file as a whole.

```{r}
birds <- add_meta(modified, birds) 
```

The built-in vocabularies are just the tip of the iceburg of established vocabularies. Here 
we add an annotation from the `skos` namespace which describes the history of where the
data comes from:

```{r}
history <- meta(property = "skos:historyNote",
  content = "Mapped from the bird.orders data in the ape package using RNeXML")
```

Because `skos` is not in the current namespace list, we add it with a url when
adding this meta element.  We also specify that this annotation be placed 
at the level of the `trees` subnode in the NeXML file.

```{r}
birds <- add_meta(history, 
                birds, 
                level = "trees",
                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
```

If we forget to add the namespace with add_meta, we can always add it later
using the `add_namespaces` function,

```{r}
birds <- add_namespaces(c(skos = "http://www.w3.org/2004/02/skos/core#"), birds)
```

(The `add_namespaces` function will avoid creating any duplicate definitions)
For finer control of the level at which a `meta` element is added, we will manipulate 
the nexml R object directly using S4 subsetting, as shown later.

<!-- Ugh. y'all have much more practice explaining this well and without tangents, 
so please feel free to edit or re-write.  -->

<!--
- Utilities to add common metadata

@@

Temporal, spatial, taxonomic coverage for dryad/knb?
-->

@@

To illustrate metadata access, we begin by reading in a NeXML file.
